---
title: "Activity Monitoring"
author: "Ramesh Balaji"
date: "1 May 2016"
output: html_document
---

Loading and processing data from Activity Data set
```{r echo=TRUE}
## loading the lattice function for displaying panel plot later in the assignment
    library(lattice)

## Working directory to be setup and activity file is loaded to a Dataframe "ActivityData"
    setwd("C:/Ramesh/R Programs/Reproducible Research/repdata-data-activity")
    ActivityData=read.csv("activity.csv",stringsAsFactors=FALSE)
    str(ActivityData)
```

Coming out Total Steps by Date based on Activity Data, find out the Mean and Median and print Histogram with Mean and Median

```{r echo=TRUE}

## Total steps taken per day
    TotalSteps <- aggregate(ActivityData$steps, by=list(date=ActivityData$date), 
                            FUN=sum, na.rm=TRUE)
    colnames(TotalSteps) <- c("Date","SumofStepsbyDate")
    TotalStepsbyDate <- as.numeric(TotalSteps$SumofStepsbyDate)

## Histogram of Total No. of Steps taken per day.
hist(TotalStepsbyDate, 
     main="Activity Data",
     xlab="Total No. of Steps", 
    border="black", 
     col="red")

## Mean of total steps taken per day
MeanSteps <- aggregate(ActivityData$steps, by=list(date=ActivityData$date),FUN=mean, na.rm=TRUE)
colnames(MeanSteps) <- c("Date","MeanbyDate")
MeanSteps$MeanbyDate

## Median of total steps taken per day
MedianSteps <- aggregate(ActivityData$steps, by=list(date=ActivityData$date), FUN=median, na.rm=TRUE)
colnames(MedianSteps) <- c("Date","MedianbyDate")
MedianSteps$MedianbyDate

```

Average Daily Activity Pattern

```{r echo=TRUE}

## Average steps taken by Interval categories of 5 minutes.
MeanStepsbyInterval <- aggregate(ActivityData$steps, 
                                 by=list(interval=ActivityData$interval), 
                                 FUN=mean, na.rm=TRUE)
colnames(MeanStepsbyInterval) <- c("Interval","AverageSteps")

## Plotting the pattern
plot(MeanStepsbyInterval$Interval, MeanStepsbyInterval$AverageSteps , 
     xlab = "Interval", 
     ylab = "AverageSteps", 
     type="l")

## finding the max of the average steps
MaxStepsbyIntervalbyDays <- max(MeanStepsbyInterval$AverageSteps,na.rm=TRUE)

## finding the interval for the max step derived in earlier step through subset function
Maxinterval <- subset(MeanStepsbyInterval,
                      AverageSteps == max(MeanStepsbyInterval$AverageSteps,
                        na.rm=TRUE))

Maxinterval

```

Input Missing Values - The Key strategy for filling in Missing values is to replace them with "Mean of Steps" present in Activity Data set.

```{r echo=TRUE}
## finding the total no. of NA rows.
SumofNARows <- sum(is.na(ActivityData))
SumofNARows

## if a specific step is "NA", then replacing it with Mean function of that step.
ActivityData$steps[is.na(ActivityData$steps)] = mean(ActivityData$steps, na.rm=TRUE)

## creating a new data frame with mean values filled in.
ActivityData_New <- ActivityData
```

Coming out Total Steps by Date based on NEW Activity Data and print Histogram with Mean and Median 

```{r echo=TRUE}
## Total steps taken per day after filling up the NA values
    TotalStepsNew <- aggregate(ActivityData_New$steps, by=list(date=ActivityData_New$date), 
                            FUN=sum, na.rm=TRUE)
    colnames(TotalStepsNew) <- c("Date","SumofStepsbyDate")
    TotalStepsbyDateNew <- as.numeric(TotalStepsNew$SumofStepsbyDate)

## Printing the new Histogram with NA values  filled in
hist(TotalStepsbyDateNew, 
     main="New Activity Data",
     xlab="Total No. of Steps", 
    border="black", 
     col="red")

## Mean of total steps taken per day after the NA values filled in
MeanStepsNew <- aggregate(ActivityData_New$steps, by=list(date=ActivityData_New$date), FUN=mean, na.rm=TRUE)
colnames(MeanStepsNew) <- c("Date","MeanbyDate")
MeanStepsNew$MeanbyDate

## Median of total steps taken per day after the NA values filled in
MedianStepsNew <- aggregate(ActivityData_New$steps, by=list(date=ActivityData_New$date), FUN=median, na.rm=TRUE)
colnames(MedianStepsNew) <- c("Date","MedianbyDate")
MedianStepsNew$MedianbyDate

```

Based on the Histogram of Total no. of steps in new activity data (NA values filled with Mean) vs old activity data where there are NA values, we see the frequency of Steps moved from 25 to 35 in the Histogram.

```{r echo=TRUE}
## Totalsteps with NA's
sum(TotalStepsbyDate)
## Totalsteps with NA's replaced by Mean steps
sum(TotalStepsbyDateNew)
## Finding out the difference in steps.
sum(TotalStepsbyDateNew) - sum(TotalStepsbyDate)
```

Finding and Printing differences in activity patterns between weekdays and weekends in a panel plot


```{r}
ActivityData_New$date <- as.Date(ActivityData_New$date)

## using weekday function to get weekdays from the date activity
ActivityData_New$DayofWeek <- weekdays.Date(ActivityData_New$date,TRUE)

## determining whether it is a weekend or weekday
ActivityData_New$WeekdayOrWeekend <- ifelse((ActivityData_New$DayofWeek == "Sat" | ActivityData_New$DayofWeek == "Sun"), "Weekend", "Weekday")

## create a new factor variable for weekdays and weekends
ActivityData_New$WeekdayOrWeekend <- as.factor(ActivityData_New$WeekdayOrWeekend)
str(ActivityData_New)

xyplot(ActivityData_New$steps~ActivityData_New$interval|ActivityData_New$WeekdayOrWeekend,
ylab="no of steps", xlab="interval", layout=(c(1,2)), type = "l")


```

